taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-09-26 paper/manuscript.Rmd

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithet such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette).

Names may also have an ambiguous resolution *discussion about taxonomic concepts here*. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

Underlying data frames can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank. These functions allow us to ask general questions such as, how many bird species are there?

@@

```{r}
## example of sci name matching to more than one id.
```

We see that insert_name matches two accepted names which the user will have to choose between.

@@

`taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name. Common names come with many of the same problems as scientific names but worse, e.g. matching to more than one accepted name is even more frequent, non-standardized formatting (like capitalization) and spelling etc. Common names are accessed via `by_common` which takes a vector of common names. The user can then resolve discrepencies.

*talk about how we deal with common names and `by_common`*

@@

@@

We've shown identifiers are useful in an a whole bunch of applications. People haven't done it, becuase it's time consuming - but now its fast!

- `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution. Instead, it fills an important gap between existing tools and typical research patterns.  In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.

- `taxadb` works from versioned snapshots of the data providers.

- All `taxadb` functions are specific to a provider, while operating consistently across different providers.

- `taxadb` does not attempt to present any unified taxonomic backbone or synthesis from existing naming providers, or make any assertions, numerical scores or other inferences about the data or matches to the data -- it is merely a tool for accessing this infromation.  This contrasts from other approaches such as the Global Names Resolver, which

There is an important objection to this approach that must also be addressed.

@Franz2018

>>  no self-respecting biodiversity researcher would or should trust aggregated data blindly, that indeed careful data cleaning is almost always necessary and expected to render the downloaded data fit for purpose, and that therefore aggregator services need to be understood mainly or merely as data discovery tools.

> We reject this deflationary view for four reasons. For one, aggregators frequently blur the lines between advertizing their services just as a data discovery tool or as a more powerful data signal tool. Second, the biases inherent in using unitary backbones remain in place even if users are only interested in discovering all relevant data for their research purpose. If the backbone-based data record modulations are not easily retrievable through primary on-line interfaces, then users are significantly constrained in their ability to design search queries with high rates of precision and recall (81). Third, let us assume that labor-intensive off-line data quality review and correction efforts are indeed the norm, prior to publishing. Then why must the fruits of these efforts remain outside of the aggregator’s environment? Why can they not immediately flow back into the same aggregation domain, while recording the provenance of expert changes? In other words, if the workflow of rendering data fit for purpose flows only in one direction, i.e. from the on-line aggregate to the off-line quality review and publication, then our criticism of the design stands.

The inability to make automate use and manipulation

(Figure 1: schematic example of data join?)

Assemble data across multiple studies

How many species of Chameleons are known? There are over

```{r}
taxa_tbl("col") %>% count(taxonomicStatus, sort = TRUE)
```

>>>>

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette).

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

Underlying data tables can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank. These functions allow us to ask general questions such as, how many bird species are there?

@@

```{r}
## example of sci name matching to more than one id.
by_name("Zosterops rendovae") %>% 
  select(1:5)

We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between.

@@

`taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name. Common names come with many of the same problems as scientific names but often more frequent, e.g. matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc. Common names are accessed via `by_common` which takes a vector of common names. The user can then resolve discrepencies.

@@

<!--
What other aspects should be included in the table? Citations?
-->

```{r cache = FALSE, echo = FALSE}
tibble(provider = c("Integrated Taxonomic Information System", 
                                 "National Center for Biological Information's Taxonomy database",
                                 "Catalogue of Life",
                                 "Global Biodiversity Information Facility Taxonomic Backbone",
                                 "FishBase",
                                 #"SeaLifeBase",
                                 "Wikidata", "Open Tree Taxonomy",
                                 "International Union for Conservation of Nature and Natural Resources"),
                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
                                     "wd", "ott", "iucn"),
                    total_identifiers = map(abbreviation, 
                                            ~taxa_tbl(.x) %>% 
                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
) %>% printtable()

@@

Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications. Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary fo resolving even moderately sized data sets. `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers. `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.

`taxadb` establishes a local database of provider allowing for consistent and reproducible queries. The set of functions operate consistently accross providers returning data in a standardized format.
